Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-02-18 paper/main.tex
Added paper ID
revised section 3 GPU implementation started section 6

\documentclass[10pt,twocolumn,letterpaper]{article}

@@

\def\iccvPaperID{****} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

@@

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}
}

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, the face
recognition system has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. Since any algorithm
parallelization is highly dependent on the specific computer hardware, the
architecture of our test system will be explained in detail in Section
\ref{sec:parallelism}.  We show that a highly accurate and robust face
recognition system can be constructed to recognize face images from a moderate
database of several hundred subjects. Note that with the possible continual
increase of the number of subject classes, it is possible to modularize and
dispatch different stages of the face recognition routine to a network of
workstations to form an even larger scale networked system. Nevertheless, this
scenario is not the focus of this paper. The source code of our algorithms,
which is based on BLAS and Nvidia CUDA libraries, will be made available on our
website to aid peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline implemented on massively parallel hardware.}
\label{fig:pipeline}
\end{figure}

@@

The ALM algorithm requires many matrix-vector operations, this gives many-core
processors such as GPUs a computational advantage for large problem sizes.
GPUs are comprised of several streaming multiprocessors (SMP) (16 for the
Nvidia Fermi) which each contains 32 floating-point units.  Each SMP contains
its own L1-cache and shared memory and all SMPs share a common L2-cache.  The
GPU executes programs as SIMT (Single Instruction Multiple Threads) in groups
of 32 threads called \emph{warps}.  These warps are then grouped into higher
levels called \emph{thread blocks} and \emph{grids}. The GPU contains an
inverted memory hierarchy so that the L1 cache stays close to the cores,
providing the increase in performance and a computational advantage over CPUs.
Unlike normal CPUs, the Nvidia Fermi has 768 kb of L1 cache/shared memory and
768kb of L2 cache. However, this increase in performance does not come for
free. As the GPU has its own memory system, any data the GPU uses must be
transferred from RAM to the GPU over PCI-Express, which has limited bandwidth.
However, the cost of these transfers are amortized by doing large amounts of
computation at once, requiring code to be run completely on the GPU.

In summary, a programmable GPU can be analogous to a multi-core CPU processor,
except with more sophisticated vector processing hardware.  Each thread block
is a virtualized multiprocessor, which each thread as a virtualized scalar
processor.  All of these differences allow GPUs to perform intensive
calculations such as matrix-vector operations at much faster rates than CPUs.

@@

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
{\bf AY: Add an algorithm here describe the recognition block in the pipeline. Explain the algorithm and necessary modification from generic l-1}

\subsection{CPU Implementation}
{\bf AY: A diagram should be used to illustrate how the jobs are distributed on a CPU architecture.}

\subsection{GPU Implementation}
{\bf AY: A diagram should be used to illustrate how the jobs are distributed on a CPU architecture.}
The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. Because data transfer from the RAM to GPU suffers bandwidth limitations, we designed our implementation to transfer the A matrix and measurement vector at once.  The algorithm is similar
to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU keeps all of the data and runs the computation.

@@

@@

>>>>

\documentclass[10pt,twocolumn,letterpaper]{article}
% ICCV Rules:
% 8 page limit (6 free + 2 paid)
% 10MB pdf file size limit
% 30MB supplementary materials file size limit (PDF or ZIP)

@@

\def\iccvPaperID{257} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

@@

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{Parallelization of Fast $\ell_1$-Minimization for Face Recognition
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}
}

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, the face
recognition system has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. Since any algorithm
parallelization is highly dependent on the specific computer hardware, the
architecture of our test system will be explained in detail in Section
\ref{sec:parallelism}.  We show that a highly accurate and robust face
recognition system can be constructed to recognize face images from a moderate
database of several hundred subjects. Note that with the possible continual
increase of the number of subject classes, it is possible to modularize and
dispatch different stages of the face recognition routine to a network of
workstations to form an even larger scale networked system. Nevertheless, this
scenario is not the focus of this paper. The source code of our algorithms,
which is based on CUBLAS and Nvidia CUDA libraries, will be made available on our
website to aid peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline implemented on massively parallel hardware.}
\label{fig:pipeline}
\end{figure}

@@

The ALM algorithm requires many matrix-vector operations, this gives many-core
processors such as GPUs a computational advantage for large problem sizes.
GPUs are comprised of several streaming multiprocessors (SMP) (15 for our
NVIDIA Fermi) which each contains 32 floating-point units.  Each SMP contains
its own L1-cache and shared memory and all SMPs share a common L2-cache.  The
GPU executes programs as SIMT (Single Instruction Multiple Threads) in groups
of 32 threads called \emph{warps}.  These warps are then grouped into higher
levels called \emph{thread blocks} and \emph{grids}.

Fermi's inverted memory hierarchy (aggregate l1 capacity is greater than l2 capacity) is designed to support massively parallel workloads in which there is more task-private data than shared data, providing an increase in performance and a computational advantage over CPUs.
For our system, each SMP has 64 kb of L1 cache with a total of 960 kb of L1 cache/shared memory and
768kb of L2 cache. However, this increase in performance does not come for
free. As the GPU has its own memory system, any data the GPU uses must be
transferred from RAM to the GPU over PCI-Express, which has limited bandwidth.
However, the cost of these transfers are amortized by doing large amounts of
computation at once, requiring code to be run completely on the GPU.  In 2011, NVIDIA is developing a new ARM-based CPU system, dubbed 
"Project Denver" to fix performance issues such as the limited bandwidth due to PCI-E \cite{Dally-Web}.

In summary, a programmable GPU can be analogous to a multi-core CPU processor,
except with more sophisticated vector processing hardware.  Each thread block
is a virtualized multiprocessor, which each thread as a virtualized scalar
processor.  All of these differences allow GPUs to perform intensive
calculations such as large scale matrix-vector operations at much faster rates than CPUs.

@@

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
{\bf AY: Add an algorithm here describe the recognition block in the pipeline. Explain the algorithm and necessary modification from generic l-1}
{\bf VS: Why did we modify the algorithm for the recognition stage? }

\begin{algorithm}[t]
\caption{\bf (Augmented Lagrange Multiplier Method for
Recognition)}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\y \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \y$, $\blamda_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}\left(\y - A\x_k +
\frac{1}{\mu}\blamda_k, \frac{1}{\mu}\right)$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}\left(\z_l +
\frac{1}{\gamma}A^T\left(\y - A\z_l - \e_{k+1} +
\frac{1}{\mu}\blamda_k\right), \frac{1}{\mu\gamma}\right)$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
\sqrt{1+4t_l^2}\right)$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\blamda_{k+1} \leftarrow \blamda_k + \mu (\y - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\label{alg:alm_rec} 
\end{algorithm}

\subsection{CPU Implementation}
{\bf AY: A diagram should be used to illustrate how the jobs are distributed on a CPU architecture.}
The CPU ALM algorithm was written in C and compiled with Intel's MKL libraries, which provide
automatic parallelization.

\subsection{GPU Implementation}
{\bf AY: A diagram should be used to illustrate how the jobs are distributed on a GPU architecture.}
The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. Because data transfer from the RAM to GPU suffers bandwidth limitations, we designed our implementation to transfer the A matrix and measurement vector at once.  The algorithm is similar
to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU keeps all of the data and runs the computation.

@@

In this section, we present four experiments to benchmark the performance of our algorithm on CPU and GPU platforms. 
To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Unless otherwise specified, all implementations utilize single
precision floating point datatypes.

The first experiment is designed to compare the two implementations of PALM
that solve a single large instance of the $\ell_1$-minimization problem (Algorithm \ref{alg:alm}).  
The measurement $d \times n$ matrix $A$ was generated as a random Gaussian matrix, with each entry
randomly generated from the standard normal and normalized with unit norm.  The ratio of $d/n$ varied from
$10:1$ to $10:5$ with the dimension of $d$ of matrix $A$ varying from 500 to 8000.
The ground truth signals, $x_0$ and $y_k$ had a sparsity rate of 0.1 with elements generated using the standard normal and normalized with unit norm.
The measurement vector was generated with the equation $\y_0 = A \x_0 + \y_k$.    We set the number of inner iterations at 50 and outer iterations at 5000.
We also varied the different inner and tolerance levels from 1e-3 to 1e-6.
We averaged the the time taken for the ALM algorithm to solve 10 instances of a problem.

The x set of experiments will compare the CPU and GPU versions of the entire pipeline, 
which includes the alignment and recognition stage.

We pl

@@

We plot the resulted of the simulation benchmark below.  The x-axis represents the size of the 
$A$ and $A'A$ matrix and the y-axis represents the average amount of time it to complete
a single problem instance.

The observations of this experiment are summarized below:
1.  With similar accuracy rates, the GPU implementation runs faster than the CPU when
